Type theories as quotient inductive-recursive types

### 2025-02-21 tex/main.tex
another pass through Sections 3--6

Substitutions from $\Gamma$ to $\Delta = \emptyctx, A_1, \ldots, A_n$ can be understood intuitively as lists of terms $t_i : A_i$ in context $\Gamma$ for each type $A_i$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which substitutes the most recently bound variable by the term~$t$:
\[
  \emptysub : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx
  \qquad\qquad
  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}) \to \Sub\;\Gamma\;(\Delta, A).
\]
Note that the type substitution $\sub{A}{\sigma}$ is needed, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions $\idS$ and compositions $\sigma;\tau$.
%because the substitution calculus is supposed to be the initial \emph{category} with families:
\[
  \idS : \implicit{\Gamma}\; \Sub\;\Gamma\;\Gamma
  \qquad\qquad \blank;\blank : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta.
\]
These need to satisfy laws, which we will introduce later as equality constructors.
Context extension is characterised by its projections, which forgets and picks out the most recently substituted term respectively:
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\sub{A}{\pi_1\,\sigma})
\end{alignat*}
%
Again, a type substitution $\sub{A}{\pi_1\,\sigma}$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in $\Delta$ to terms in $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;\Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
\end{alignat*}
%Note that $\sub{\blank}{\blank}$ is overloaded for type and term substitution.

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Ty\;\Gamma} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIR definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%

@@

The lifting $\sigma^+$ interacts well with identities and composition.
For comparision, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
%
\begin{restatable}{lemma}{lemIdentities}\label{lem:lifting}
  For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
\end{restatable}

@@

We can add extensional identity types, a hierarchy of Coquand universes~\cite{Coquand2018}, and a unit type with an eta law, all without any additional difficulty. In particular, all these additions can be done without introducing any transports in the syntax. However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$ and its elimination principle:
%
\begin{alignat*}{3}
  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}}) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}}) \\
  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{b}})
\end{alignat*}
%
where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ is the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place. We can add the expected computation rules 
%
%\begin{alignat*}{5}
%  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
%  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
%\end{alignat*}
%
as constructors, e.g.\ $ \bool\beta_{\bfalse} : \implicit{\Gamma, P, f, t}\; \belim\,P\,f\,t\,\bfalse = f$, and the structural rule for type substitution as another straightforward clause
$\subTy{\bool}{\sigma} \reduce \bool$,
but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following
%
\begin{alignat*}{5}
  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
           & =^{\Tm\,\Gamma}_{\alert{p\,b}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\bfalse}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\btrue}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
\end{alignat*}
where the proof $p\,x : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types. Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems. Hence, we do not know how to get rid of these transports.

>>>>

Substitutions from $\Gamma$ to $\Delta = \emptyctx, A_1, \ldots, A_n$ can be understood intuitively as lists of terms $t_i : A_i$ in context $\Gamma$ for each type $A_i$ in $\Delta$: we have the empty substitution $\emptysub$, and substitution extensions $\sigma, t$ which substitutes the most recently bound variable by the term~$t$:
\[
  \emptysub : \implicit{\Gamma}\;\Sub\;\Gamma\;\emptyctx
  \qquad\qquad
  \blank,\blank : \implicit{\Gamma, \Delta, A}\; (\sigma : \Sub\,\Gamma\,\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}) \to \Sub\;\Gamma\;(\Delta, A).
\]
Note that the type substitution $\sub{A}{\sigma}$ is needed in the type of $\blank,\blank$, because $A$ is well-formed in context $\Delta$ rather than in $\Gamma$.
We also ask for identity substitutions and composition $\sigma;\tau$.
%because the substitution calculus is supposed to be the initial \emph{category} with families:
\[
  \idS : \implicit{\Gamma}\; \Sub\;\Gamma\;\Gamma
  \qquad\qquad \blank;\blank : \implicit{\Gamma, \Delta, \Theta}\;\Sub\;\Gamma\;\Delta \to \Sub\;\Delta\;\Theta \to \Sub\;\Gamma\;\Theta.
\]
These need to satisfy laws, which we will introduce later as equality constructors.
Context extension is characterised by its projections, which forgets and picks out the most recently substituted term respectively:
%Context comprehension is given by projections from \emph{non-empty} substitutions $\sigma : \Sub\;\Gamma\;(\Delta, A)$ for its \emph{tail} and its \emph{head}:
\begin{alignat*}{3}
  \pi_1 & : \implicit{\Gamma, \Delta, A}\;\Sub\;\Gamma\;(\Delta, A) \to \Sub\;\Gamma\;\Delta \\
  \pi_2 & : \implicit{\Gamma, \Delta, A}\;(\sigma : \Sub\;\Gamma\;(\Delta, A)) \to \Tm\;\Gamma\;(\sub{A}{\pi_1\,\sigma})
\end{alignat*}
%
Again, a type substitution $\sub{A}{\pi_1\,\sigma}$ is needed for~$\pi_2$, because $A$ is well-formed in context $\Delta$ instead of $\Gamma$.
Finally, we have term substitution, which given $\sigma : \Sub\;\Gamma\;\Delta$ turns terms in $\Delta$ to terms in $\Gamma$, appropriately preserving the type, which is also substituted:
\begin{alignat*}{3}
  \sub{\blank}{\blank} & : \implicit{\Gamma,\Delta, A}\;\Tm\;\Delta\;A \to (\sigma : \Sub\;\Gamma\;\Delta) \to \Tm\;\Gamma\;(\sub{A}{\sigma}).
\end{alignat*}
%Note that $\sub{\blank}{\blank}$ is overloaded for type and term substitution.

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend the substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
In a QIIT definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && \sub{(\El\, u)}{\sigma} =^{\Ty\;\Gamma} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}\sub{u}{\sigma}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $\sub{(\El\,u)}{\sigma}$ is, informally, the type of the term substitution $\sub{u}{\sigma}$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $\sub{u}{\sigma} : \Tm\;\Gamma\;(\sub{\UU}{\sigma})$ into $\Tm\;\Gamma\;\UU$.
%
In our QIIRT definition, $\transfib{\Tm\;\Gamma}{[]\UU}{\blank}$ is no longer needed, since $[]\UU$ becomes strict. 
Likewise, the equality constructor $[]\El$ may be replaced by the reduction
\begin{alignat*}{3}
  \subTy{(\El\, u)}{\sigma} & \reduce \El\,(\sub{u}{\sigma})
\end{alignat*}
Yet, this naive change breaks local confluence!
The problem is that there is now additional interaction between types and terms.
% Observe that
The local peak
\(
  \El\,u \Leftarrow \subTy{(\El\,u)}{\idS} \reduce \El(\sub{u}{\idS})
\)
cannot reduce to the same term, since the term substitution $\sub{u}{\idS}$ is merely a constructor.
%

@@

The lifting $\sigma^+$ interacts well with identities and composition.
For comparison, the reader is invited to prove the corresponding QIIT version with transports of the following lemma.
%
\begin{restatable}{lemma}{lemIdentities}\label{lem:lifting}
  For all contexts $\Gamma$, $\Delta$, $\Theta$, types $A : \Ty\,\Theta$ and substitutions $\sigma : \Sub\,\Gamma\,\Delta$ and $\tau : \Sub\,\Delta\,\Theta$, we have $\idS^+ =^{\Sub\;(\Theta, A)\;(\Theta, A)} \idS$ and $(\sigma; \tau)^+     =^{\Sub\;(\Gamma, \subTy{A}{\sigma; \tau})\;(\Theta, A)} \sigma^+ ; \tau^+$.
\end{restatable}

@@

We can add extensional identity types, a hierarchy of Coquand universes~\cite{Coquand2018}, and a unit type with an eta law, all without any additional difficulty. In particular, all these additions can be done without introducing any transports in the syntax. However, adding an inductive type such as the Booleans does introduce a little bit of friction; let us see why.
We introduce the type $\bool : \implicit{\Gamma}\;\Ty\;\Gamma$ itself, its constructors $\btrue, \bfalse : \implicit{\Gamma}\;\Tm\;\Gamma\;\bool$, and its elimination principle:
%
\begin{alignat*}{3}
  \belim &: \implicit{\Gamma} \; && (P : \Ty\;(\Gamma , \bool)) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}}) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}}) \\
  &&& \qquad\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad \to (b : \Tm\;\Gamma\;\bool) \to \Tm\;\Gamma\;(\subTy{P}{\instSub{b}})
\end{alignat*}
%
where $\instSub{t} \defeq (\idS, t) : \Sub\;\Gamma\;(\Gamma, A)$ is the substitution that substitutes $t : A$ for the top variable and leaves all other variables in place. We can add the expected computation rules 
%
%\begin{alignat*}{5}
%  \bool\beta_{\bfalse}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\bfalse && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\bfalse}})} {} && f \\
%  \bool\beta_{\btrue}           & : \implicit{\Gamma, P, f, t}\; && \belim\,P\,f\,t\,\btrue && =^{\Tm\;\Gamma\;(\subTy{P}{\instSub{\btrue}})} {} && t
%\end{alignat*}
%
as constructors, e.g.\ $ \bool\beta_{\bfalse} : \implicit{\Gamma, P, f, t}\; \belim\,P\,f\,t\,\bfalse = f$, and the structural rule for type substitution as another straightforward clause
$\subTy{\bool}{\sigma} \reduce \bool$,
but when we try to add the structural rule for term substitution applied to the eliminator, we end up with the following
%
\begin{alignat*}{5}
  []\belim & : \implicit{\Gamma, \sigma, P, f, t, b}\; \sub{(\belim\,P\,f\,t\,b)}{\sigma} \\
           & =^{\Tm\,\Gamma}_{\alert{p\,b}} \belim\,(\sub{P}{\sigma^+})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\bfalse}{\textcolor{black}{\sub{f}{\sigma}}}})\;(\alert{\transfib{\Tm\,\Gamma}{p\,\btrue}{\textcolor{black}{\sub{t}{\sigma}}}})\; (\sub{b}{\sigma})
\end{alignat*}
where the proof $p\,x : \subTy{P}{\instSub{\sub{x}{\sigma}} ; \sigma^+} = \subTy{P}{\sigma ; \instSub{x}}$ is needed to line up the types. Even though this is an equation between type substitutions, which we could hope make compute, there is no obvious way to direct the equation, and simply adding it as a reduction leads to confluence problems. Hence, we do not know how to get rid of these transports.
